---
title: "Untitled"
author: "hannah"
date: "1/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(here)
library(readr)
```

## about the data

this data is from https://aca6.accela.com/CALCANNABIS/Cap/CapHome.aspx?module=Licenses
on cannabis cultivation licences 
see also https://cannabis.ca.gov/check_a_license/

```{r}

cult <- here("cannabis/data/ReocrdList20190123.csv") %>% read_csv()

```

clean up the data a bit.

check that the last column is empty, and if so, dump it.

```{r}
cult %>% count(X9) # yes, all NAs

cult <- cult %>% select(-X9)
```

now let's clean up the column names a bit. R will let you have spaces in your column names, but the spaces can be a pain to work with. 

```{r}
cult %>% colnames()
cult2 <- cult %>% 
  # do this one manually
  rename(City_County=`City - County`) %>% 
  # for the rest, just replace any spaces in the column name with an underscore
  rename_all(funs(stringr::str_replace_all(.," ","_")))
cult2 %>% colnames()
```


Now let's explore the data.

Is the license number a unique record identifier?

```{r}
# how many rows are there in our data?
cult2 %>% nrow() # 9472
# how many unique values are there of License_Number?
cult2 %>% select(License_Number) %>% n_distinct() # 9463
```

It looks like there are a few duplicate license numbers. Let's look at them.

```{r}
cult2 %>% group_by(License_Number) %>% filter(n()>1)

```

they are exact duplicates on all columns. We can confirm that by seeing how many unique rows there are.

```{r}
cult2 %>% n_distinct() # 9463, same as the number of unique license numbers.
```

Remove the duplicates.

```{r}
cult3 <- cult2 %>% distinct()
cult3 %>% nrow() # 9463, as expected. good.
```

Let's split out city and county

```{r}
library(stringr)
cult3 <- cult3 %>% 
  mutate(
    county = ifelse(str_detect(City_County," - "), word(City_County, start=2, sep=" - "), City_County),
    city   = ifelse(str_detect(City_County," - "), word(City_County, start=1, sep=" - "), NA)
  )
```


What types of licenses are there?

```{r}
cult3 %>% count(License_Type, sort=TRUE)
cult3 %>% count(Type_of_License, sort=TRUE)

```

What counties have the most licences?

```{r}
cult3 %>% count(county, sort=TRUE)

```

Who are the big players?

```{r}
cult3 %>% count(Legal_Business_Name, sort=TRUE)
```


