---
title: "Untitled"
author: "Sandhya"
date: "1/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

install.packages("rvest")
install.packages("stringr")
install.packages("tidyr")

library(here)
library(rvest)
library(stringr)
library(tidyr)

```

## About the data

Rvest is used for scraping data from the web and includes commands html and html_node
Stringr is useful for working with strings, including matching, subsetting and extracting data
Tidyr is a common package used for reshaping data

One row per superbowl 

we want to look at the superbowl data over time 

#First assign the web address to url
```{r}
url <- 'http://espn.go.com/nfl/superbowl/history/winners'

```
Pass it through the command called 'read_html'
read_html is part of the rvest package and pulls the underlying html from a webpage.

```{r}
superbowl <- read_html(url)
  # Now we will use the rvest functions html_nodes and html_table 
  # We'll extract the HTML table and convert it to a data frame
  superbowl_table <- html_nodes(superbowl, 'table')
  sb <- html_table(superbowl_table)[[1]]
  # Check the data
  head(sb)
  #What should we remove from the table?
  sb <- sb[-(1:2), ]
  #Check the data
  head(sb)
  #Now let's set new column names or variables
  names(sb) <- c("number", "date", "site", "result")
  #Check the data
  head(sb)

```


```{r}
  #We can also clean up the data here.
  #We'll replace Roman numerals with numeric values
  sb$number <- 1:51
  #Check the data
  head(sb)
  #We will also convert the date to a standard format
  sb$date <- as.Date(sb$date, "%B. %d, %Y")
  #Check the data
  head(sb)
  #The data column should be split so we can more easily work with the data and filter it. 
  #We'll divide it into 3 columns. 
  sb <- separate(sb, date, c('year', 'month','day'),sep = "-")
  #Check the data
  head(sb)
  #The result column should be split so we can more easily work with the data.
  #We'll divide it into 4 new columns. Let's start by splitting by the comma delimiter.
  sb <- separate(sb, result, c('winner', 'loser'), sep=', ', remove=TRUE)
  #Check the data
  head(sb)
```


Now we'll split out the scores from the winner and loser columns.
We'll do this by substring pattern matches, which is based on regex.
Let's look at the table:


```{r}
scorepattern <- " \\d+$"
sb$winnerScore <- as.numeric(str_extract(sb$winner, scorepattern))
sb$loserScore <- as.numeric(str_extract(sb$loser, scorepattern))
sb$winner <- gsub(scorepattern, "", sb$winner)
sb$loser <- gsub(scorepattern, "", sb$loser)
```
Let's write the result out to a csv. 
write.csv(sb, 'superbowl.csv', row.names=F)
View(sb)

Let's check for duplicates.

R has a very useful function for detecting duplicate values
```{r}
duplicated(sb)
```
If there are duplicates, we can remove the duplicate values.

```{r}
sb <-unique(sb)
#Now, let's sort our data. 
sort(sb$winner)
```